\thetitle

### 2020-07-06 writing/challenges.tex
draft of rational programmer

Besides the blame information they produce, the three gradual typing
systems also differ in the runtime type checks they perform. Therefore an
application of the evaluation method to a gradual typing system has to
factor in the effect of type checks to determine whether blame helps the
rational programmer find an impedance mismatch. A solution to this issue
is yet another mode for the rational programmer that uses information only
from teh underlying languages exceptions to select the next component(s) to type.
This \emph{vanilla} mode simulates a rational programmer that ignores
blame and serves as the baseline for blame's value.

The final big challenge is that the different gradual type
systems, the number of necessary modes for the rational programmer
and the large number of mutants lead to an explosion of scenarios the
method needs to examine. Thus a practical application of the
method has no option but to carefully sample this space while 
making sure that it still produces reproducible conclusions.

@@

\item Implement all the gradual type systems to be compared on a common
  basis (section~\ref{sec:landscape}).

\item Construct a large number of suitable benchmarks for all gradual type
  systems with representative
  type errors  (section~\ref{sec:mutate}).

\item Simulate the rational programmer with sufficient 
  modes to (i) match the peculiarities of each gradual type system and 
    (ii) isolate blame from other confounding factors
    (section~\ref{sec:rational}).

>>>>

Besides the blame information they produce, the three gradual typing
systems also differ in the runtime type checks they perform. Therefore an
application of the evaluation method to a gradual typing system has to
factor in the effect of type checks to determine whether blame helps the
rational programmer find the cause of an impedance mismatch. A solution to this issue
is yet another mode for the rational programmer that uses information only
from the underlying languages exceptions to select the next component(s) to type.
This \emph{exception} mode simulates a rational programmer that ignores
blame and serves as the baseline for blame's value.

The final big challenge is that the different gradual type
systems, the number of necessary modes for the rational programmer
and the large number of mutants lead to an explosion of scenarios the
method needs to examine. Thus a practical application of the
method has no option but to carefully sample the space of scenarios while 
making sure that it still produces reproducible conclusions.

@@

\item Implement all three gradual type systems on a common
  basis (section~\ref{sec:landscape});

\item Construct a large number of suitable benchmarks for all gradual type
  systems with representative
  type errors  (section~\ref{sec:mutate});

\item Simulate the rational programmer with sufficient 
  modes to (i) match the peculiarities of each gradual type system and 
    (ii) isolate blame from other confounding factors
    (section~\ref{sec:rational});

### 2020-07-06 writing/rational.tex
draft of rational programmer

Consider a rational programmer  that tries to debug a faulty program in
Typed Racket with the default natural semantics. If the Typed Racekt
typechecker detects the issue with the program, the programmer knows where
to look for the bug; in the component that doesn't type check. However, if
the type checker blesses the program, the programmer relies on  runtime
type checks  to  signal an impedance mismatch between the
program's behavior and its types. In these checks raise an error, the programmer ``follows'' blame to
locate where to add a type specification next.  Hopefully, the extra
types are sufficient to turn the run time error into a static one. Otherwise,
the rational programmer starts over with the latest version of the program. 
In essence, at each iteration the
faulty program migrates to a version with more type specifications until
the faulty component becomes typed.

Therefore, a concise way to model rational programmers is with the
migratory paths they follow through the \emph{configuration space} of the
program. This space consists of variants of the program, dubbed
\emph{configurations}, that differ in their type specifications. In the 
setting of Typed Racket's migratory typing,\footnote{The
method can be adapted to so-called true gradual typing
systems~\cite{svcb-snapl-2015} in a straightforward manner.}, each
configuration differs from another in the components that have type
specifications. Hence, formally, we can represent a program
$\system$ as a set of components $\set{\component}$ and each
configuration $\conf$  as a subset of $\set{\component}$ that contains
exactly the components of  $\system$ that are typed in $\conf$. Naturally
the configurations of  $\system$ are ordered by the subset relation and
thus form a lattice $\lattice{\system}$ with $2^{\size{\system}}$ elements.
The bottom configuration of $\lattice{\system}$ is always $\emptyset$ and
the top element is $\system$ itself. In between there are all the
mixed-type configurations. If we assume that  component
$\faulty{\system}$ contains the fault, the type checker detects any
configuration $\conf$ that includes $\faulty{\system}$. Any other
configuration $\conf$ of $\system$ may blame a $\component_b$, which we denote with
$\blame{\system,\conf}$.\footnote{Blame can result either directly from a Typed
Racket type check or from a check in a Racket library function that raises
a blame error or, in the absence of blame, from an
exception that we track back to the component that causes it with the help
of Racket's stack trace.}

With these definitions, the actions of the rational programmer map to an ascending chain in
$\lattice{\system}$, dubbed a \emph{blametrail}. 
The starting configuration $\conf_0$ of a blametrail,  dubbed the \emph{root of the
blametrail}, is the configuration of
$\program$ that the programmer attempts to debug. Any other
configuration $\conf_{i+1}$ in the blametrail differs from its predecessor
$\conf_{i}$ in that 
the untyped component $\blame{\program, \conf_i}$ of $\conf_{i}$
 becomes typed in $\conf_{i+1}$. Putting these pieces together:

\begin{quote}
\it 
A natural blametrail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \{\blame{\system, \conf_i}\}$.
\end{quote}

So far the definition of blametrail does not capture whether blame helps the rational
programmer locate the bug in the root of a blametrail. For that we need 
two properties of blametrails. A blametrail is \emph{final} if
the rational programmer cannot extend it further. Or formally, if 
$\blame{\program, \conf_n}$ is not defined where  $\conf_n$ is the last
configuration of the blametrail. A blametrail is \emph{successful} 
if it is final and  its last configuration
$\conf_n$ contains the faulty component, that is $\faulty{\system}
\in \conf_n$.  We can now give a precise definition of what it
means for blame to help the rational programmer in natural Typed Racket:
\begin{quote}
\it
  Natural blame is useful for a program $\system$ iff 
  all natural blametrails in  $\lattice{\system}$ are successful.
\end{quote}

However, the definition does not give a full picture of how much blame helps the
natural rational programmer. In particular, all blametrails in $\lattice{\system}$
may  be successful but this may well be because of some other aspect of
natural, such as the type checks it performs, or it may even be a result
of pure chance. To exclude these possibilities we define two different
modes for the rational programmer. In the first, the natural programmer
ignores blame and instead decides what component to type next based on
information from the stack trace of the failed runtime type check.  This 
vanilla rational programmer corresponds to \emph{vanilla natural
blametrails}:
\begin{quote}
\it 
A vanlla natural blametrail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \{\error{\system, \conf_i}\}$.
\end{quote}
where $\error{\system, \conf}$ denotes the component that triggers a
failed runtime check according to Racket's stack trace. 
In the second mode, the natural programmer
ignores any kind of information related to the type error and randomly
picks which component to type next.  This 
random rational programmer corresponds to \emph{random natural
blametrails}:
\begin{quote}
\it 
A vanlla natural blametrail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$.
\end{quote}
where $\random{\system, \conf}$ denotes a randomly selected of $\system$
that is not in $\conf_i$.

With vanilla and random blametrails, we can define the value that blame
adds to natural Typed Rachet: 
\begin{quote}
\it
  Natural blame  adds value to natural for a program $\system$ iff 
 iff 
  all natural blametrails in  $\lattice{\system}$ are successful while 
not all vanilla and random natural blametrails are successful.
\end{quote}

\noindent
{\bf Experimental Process.}~Given a set of faulty benchmarks such as the ones
from~section \ref{sec:mutate}, it is straightforward to turn the above definition
into the falsfiable hypothesis of an experiment that tests whether blame
add value to natural Typed Racket. Specifically, for each benchmark
we first construct its configuration lattice. Then, we determine the
configurations of the lattice that detect the fault in the
program. This can happen either because a
 runtime type check fails or Typed Racket's typechecker
rejects the configuration. In either case,  each of these configurations is the
root of a natural blametrail in the configuration lattice of the benchmark. \emph{We
hypothesize that all such natural blametrails are successful while the
vanilla and random ones are not}. 
To confirm or falsify the hypothesis for a benchmark, we simply extend each blametrail
 according to the mode of the corresponding rational programmer  until the blametrail
becomes final\footnote{All blametrails are final since the lattices are
finite.} and we check if its last configuration contains the faulty
component of the benchmark.

\subsection{How to Generalize the Rational Programmer Beyond Natural Typed
Racket}

Adapting the simulated rational programmer to handle transient
and erasure Typed Racket requires taking into account the distinguishing
features of each semantics. As we discuss in section~\ref{sec:landscape}, 
transient blames a set of components in contrast to natural's
single-component blame. At the other end of the blame spectrum, erasure
blames no component and relies on Racket's exceptions.

Hence, for an erasure rational programmer only the vanilla and random 
rational programmers are meaningful. In contrast, for transient, we 
need two extra transient-specific modes for the rational
programmer. Specifically, in the first mode, the rational programmer types
all blamed components.  In the second mode, the rational programmer types
the component that is added to the blame set first --- and thus can 
detect an impendance mismatch earlier in the evaluation of a program. 
As before, the two modes give
rise to two different notions of blametrail:
\begin{quote}
\it 
A  transient blametrail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
~$\conf_{i+1} \setminus \conf_i \in \{\mblame{\system, \conf_i}\}$;
\end{quote}
\noindent
and
\begin{quote}
\it 
A firstborn transient blametrail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
  $\conf_{i+1} \setminus \conf_i = \oldest{\mblame{\system, \conf_i}}$
\end{quote}
\noindent
where $\mblame{\system, \conf}$ denotes the set of components that $\conf$
blames.

\noindent
{\bf Experimental Process.} The following table summarizes the modes for the rational programmer in
different gradual typing systems:

\center
\begin{tabular}[c]{c|c|c|c}
                       & {\bf Natural}  & {\bf Transient} & {\bf Erasure}  \\
\hline 
{\bf Blame}            &       X        &                 &                \\
  {\bf Multi blame}    &                &        X        &                \\
 {\bf Firstborn blame} &                &        X        &                \\
{\bf Vanilla}          &       X        &        X        &       X        \\
{\bf Random}           &       X        &        X        &       X        \\
\end{tabular}

>>>>

Consider a rational programmer  that tries to debug a faulty program in
natural Typed Racket. If the Typed Racket
type checker detects the problem, the programmer knows where
to look for the bug; in the component that contains the type specification
the type checker complains about. However, if
the type checker blesses the program, 
the programmer relies on  runtime type checks  to  signal the impedance mismatch between the
program's behavior and its types. If these checks do raise an error that
blames a component in the program, 
the rational programmer ``follows'' blame to
locate where to add a type specification next.  Hopefully, the extra
types are sufficient to turn the run time error into a static one. Otherwise,
the rational programmer starts over with the latest version of the program. 
In other words, at each iteration the
faulty program migrates to a version with more type specifications until
it becomes ill-typed.

Therefore, a concise way to model rational programmers is with the
migratory paths they follow through the \emph{configuration space} of a
program. This space consists of variants of the program, dubbed
\emph{configurations}, that differ in their type specifications. In the 
setting of Typed Racket's migratory typing,\footnote{The
method can be adapted to so-called true gradual typing
systems~\cite{svcb-snapl-2015} in a straightforward manner.} each
configuration differs from another in the components that have type
specifications. Hence, formally, we can represent a program
$\system$ as a set of components $\set{\component}$ and each
configuration $\conf$  as a subset of $\set{\component}$ that contains
exactly the components of  $\system$ that are typed in $\conf$. Naturally
the configurations of  $\system$ are ordered by the subset relation and
thus form a lattice $\lattice{\system}$ with $2^{\size{\system}}$ elements.
The bottom configuration of $\lattice{\system}$ is always $\emptyset$ and
the top element is $\system$ itself. In between there are all the
mixed-type configurations.

With these definitions, the actions of the rational programmer map to an ascending chain in
$\lattice{\system}$, dubbed a \emph{trail}. 
The starting configuration $\conf_0$ of a trail,  dubbed the \emph{root of the
trail}, is the configuration that the programmer attempts to debug. Any other
configuration $\conf_{i+1}$ in the trail differs from its predecessor
$\conf_{i}$ in that 
the untyped component  that $\conf_i$ blames under the natural semantics, dubbed $\blame{\system, \conf_i}$,
 becomes typed in $\conf_{i+1}$.\footnote{Blame can result either directly
 from a run time type check or, in the absence of blame, from an
 exception that we track back to the component that causes it with the
 help of the stack trace.} Putting these pieces together:

\begin{quote}
\it 
A natural blame trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \{\blame{\system, \conf_i}\}$.
\end{quote}

To define capture whether blame helps rational
programmers in their search for bugs, we define when a trail is successful:
\begin{quote}
\it 
A trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful}  iff its last configuration $\conf_n$ does not type check.
  \end{quote}
  In juxtaposition:
\begin{quote}
\it 
A trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{failing}  iff $\conf_n$ type checks and $\blame{\system,\conf_n}$ is
undefined, meaning $\conf_n$ does not result in blame. 
 \end{quote}
That is failing trails are those that we cannot extend
further following blame and have not reached a configuration that reveals
the bug at
compile time.

 We can now give a precise definition of what it
means for blame to help the rational programmer:
\begin{quote}
\it
  Natural blame is useful for a program $\system$ iff 
  $\lattice{\system}$ contains no failing natural blame trails.
\end{quote}

However, this definition does not give a full picture of how much blame helps the
natural rational programmer. In particular,  a $\lattice{\system}$
may contain no failing natural blame trails because
of the run time type checks natural performs rather than blame, or even as a result
of pure chance. To account for these possibilities, we define two extra
modes for the rational programmer. In the first, the natural programmer
ignores blame and instead decides what component to type next based on
information from the stack trace of the failed runtime type check of a
configuration.  This exception mode of the rational programmer corresponds to \emph{natural
exception trails}:
\begin{quote}
\it 
A natural exception trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \{\error{\system, \conf_i}\}$
  where $\error{\system, \conf}$ denotes the component that triggers a
an exception of the underlying language.
\end{quote}

In the second mode, the natural programmer
ignores any kind of information related to the type error and randomly
picks which component to type next:
\begin{quote}
\it 
A natural random trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$
where $\random{\system, \conf}$ denotes a randomly selected component of $\system$
that is not in $\conf_i$.
\end{quote}
\noindent

With exception and random trails, we can finaly define the value that blame
adds to the natural semantics: 
\begin{quote}
\it
  Blame adds value to natural for a program $\system$ iff 
  no natural blame trails in  $\lattice{\system}$ are failing while 
 there exist at least one vanilla and one random natural trail that is
  failing.
\end{quote}

\noindent
{\bf Experimental Process.}~Given a set of faulty benchmarks such as the
ones from section~\ref{sec:mutate}, it is straightforward to turn the
above definition into the an experimental question for natural Typed Racket. 
Specifically, for
each benchmark we first construct its configuration lattice. Then, we
determine the configurations of the lattice that detect the fault in the
program. This can happen either because a runtime type check fails or
because Typed Racket's type checker rejects the configuration. In either
case,  each of these configurations is the root of a natural blame trail
in the configuration lattice of the benchmark. We ask whether
\emph{no such natural blame trail is failing}.  Furthermore  we also
examine whether \emph{there are failing exception and random trails that start
from the same roots}.  To answer the questions for a
benchmark, we simply start from our set of roots and extend each
trail according to the corresponding mode of the rational programmer
until no configurations can be added to the trail\footnote{All trails
reach that point since the lattices are finite.}. Finally we check if the
last configuration of each trail type checks or not. If it does the trail
is successful, otherwise it is failing.

\subsection{How to Generalize the Rational Programmer Beyond Natural}

Adapting the simulation of rational programmer to transient
and erasure requires taking into account the distinguishing
features of each semantics. As we discuss in section~\ref{sec:landscape}, 
transient blames a set of components in contrast to natural's
single-component blame. At the other end of the blame spectrum, erasure
blames no component and relies on the exceptions of the underlying
language.

Hence, for an erasure rational programmer only the exception and random 
rational programmers are meaningful. In contrast, for transient, we 
need two transient-specific modes for the rational
programmer instead of natural's blame mode. Specifically, in the first mode, the rational programmer types
all blamed components.  In the second mode, the rational programmer types
the component that is added to the blame set first --- and thus can 
detect an impedance mismatch earlier in the evaluation of a program. 
As before, the two modes give
rise to two different notions of trail:
\begin{quote}
\it 
A  transient all blame trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
~$\conf_{i+1} \setminus \conf_i \in \mblame{\system, \conf_i}$
where $\mblame{\system, \conf}$ denotes the set of components that $\conf$
blames under transient semantics; 
\end{quote}
\noindent
and
\begin{quote}
\it 
A transient firstborn  trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
  $\conf_{i+1} \setminus \conf_i = \first{\mblame{\system, \conf_i}}$
  where $\first{\mblame{\system, \conf}}$ denotes the first component 
  the transient semantics add to the blame set for $\conf$.
\end{quote}

\noindent
{\bf Experimental Process.} Similar to the experiment for natural, we can
test whether blame adds value to transient. However, we must ask two
separate questions about blame; one for each mode of the transient
rational programmer. Specifically, we have two derivative questions: (i)
does blame add value to transient if the rational programmer focuses on all
blamed components of a configuration and (ii) does blame adds value to
transient if the rational programmer focuses on the earliest entry in a
configuration's blame set? The two questions boil down to using 
transient all blame trails and transient
firstborn trails respectively and performing twice the same experiment as that
for natural blame.

In the case of erasure, instead of blame, we can examine
the value of runtime exceptions for the rational programmer. Thus, in the
same vain as for transient and natural we
ask whether the configuration lattice of a given program 
contains only successful erasure exception trail and at least one failing
erasure random trail. Of course, we can formulate the same questions about 
 exceptions for natural and transient and query the value of
exceptions for these semantics too.

Overall our experiment examines six questions for each benchmark: one for the value of natural
blame, two for the value of transient blame and one for the value of
exceptions in each of natural, transient and erasure.
Table~\ref{fig:experiment-outline} summarizes how each question relates to different
trails/modes of the rational programmer. For example 
experimental question $Q_1$ asks whether blame adds value to transient and 
our experiment uses the natural blame, exception and random trails to
answer it.

\begin{figure}
\center
{\begin{tabular}[c]{c|c|c|c}
                        & {\bf Natural}        & {\bf Transient}          & {\bf Erasure}  \\
\hline 
{\bf Blame}             &       $Q_1$          &                          &               \\
{\bf All blame}         &                      &     $Q_2$                &               \\
{\bf Firstborn blame}   &                      &     $Q_3$                &               \\
{\bf Exception}         &       $Q_1$/$Q_4$    &     $Q_2$/$Q_3$/$Q_5$    &      $Q_6$    \\
{\bf Random}            &       $Q_1$/$Q_4$    &     $Q_2$/$Q_3$/$Q_5$    &      $Q_6$    \\
\end{tabular}}
  \caption{ Experimental questions and rational programmer modes. $Q_1$:
  Does natural blame add value to natural? $Q_2$: Does transient all blame
  add value to transient?  $Q_3$: Does transient firstborn blame add value
  to transient?  $Q_4$: Do exceptions add value to natural?  $Q_5$: Do
  exceptions add value to transient? $Q_6$: Do exceptions add value to
  erasure? } 
  \label{fig:experiment-outline}
\end{figure}

\noindent
{\bf Programmer Effort.} In addition to the value of blame, trails permit
us to examine the effort a rational programmer spends to locate a bug. 
For instance, the length of a successful natural blame trail tells us how
many components the programmer has to type before reaching the faulty
component. This idea also applies to all other modes of the rational
programmer. The most interesting case is that for transient all blame trails. 
In detail, given a successful transient all blame trail 
$\conf_0, ...., \conf_n$ in a configuration lattice $\lattice{\system}$,  
the effort of the rational programmer is the 
number of distinct configurations in the all blame trails with root $\conf_n$ and 
length $n$ in $\lattice{\system}$. Our experiment utilizes this notion of
effort to analyze burden on the rational programmer for finding a
bug when the answer to one of the experimental questions from above is
positive for a benchmark.
